The ChEMBL database as Linked Open Data

### 2012-08-20 article.tex
edit discussion that was a little clumsy and possibly more negative than needed

For example, much can be argued about the choice of ontologies, and the lack of an
ontology at places where we decided to use custom classes and predicates. Indeed, the
choices have only partly been decided on ontological arguments, but are mostly based
on pragmatic approaches to fulfill our needs for the applications. It should be noted,
however, that the ChEMBL data base captures a wide variety of experimental studies
and that semantics are not fully specified in the source from which the data was
extracted. Clearly, missing semantics must not be made up: it is better to leave
some semantics undefined than making incorrect assumptions. Indeed, during the
development of the applications some incorrect assumptions have been discovered and
fixed. Therefore, the goal of this work is the conversion of the ChEMBL data into
RDF, and not an ontological modeling of the domain.

For example, this applies for the links to other databases: when linking between
two databases based on a shared identifier, we make assumptions. Besides the mere
aspect that either database may contain typing mistakes in databases, additional
error may be introduced because at least the assumption is made in the ChEMBL database
that entries in that database really matches the data linked to in the other database.
That is the reason why we used different predicates for linking to other databases.
This problem, however, is general, and applies to other methods to link databases with
identifiers too~\cite{VanIersel2010}.

Also debatable is the choice of the ontologies. For example, in this release of the
ChEMBL-RDF data, we adopted the cheminformatics ontology, CHEMINF. Using this ontology
we can encode numeric and non-numeric descriptors of chemical entities in the same
standardized way. Using this approach we expose identity descriptors such
as the SMILES string, InChI, and InChIKey, and molecular properties including the
logP, rotatable bond count, and hydrogen bond donor and acceptor counts.
The design of this annotation is more complex than strictly needed for a translation
of the data in the ChEMBL relational database into RDF. Moreover, this also
affects the querying of the data, where a more complex SPARQL query is needed.
While we found that our selection of ontologies fit well with our applications,
we anticipate that we will select more expressive ontologies in the future.

For example, not all information in the databases has so far been annotated using
standard ontologies. That includes the activity types (``IC50'', ``EC50'', and ``Potency'' in ChEMBL), making it
still impossible to reason or group over similar activity types. For example,
the IC$_{50}$ and K$_i$ end points are closely related, but that is currently not
ontologically expressed. Similarly, measurement units are currently provided as
string literals, making it needlessly hard to compare concentrations in nM and mM.
If we further allow to make some assumptions, we can even search for activities
expressed in $\mu$g/ml with query concentrations in nM. Without introducing machine-readable
semantics, such conversions are less easily automated.

To address these issues, two ontologies were
recently developed. First, the BioAssay Ontology addresses many of our
remaining concerns around making the experimental data machine readable~\cite{Visser2011}. Second,
the Quantity, Unit, Dimension and Type (QUDT) ontology makes it possible
to express measurements in a machine readable way~\cite{QUDT}. Both approaches are
currently explored to be applied to ChEMBL-RDF as part of work in the
Open PHACTS project~\cite{Williams2012}.

Taking these points into account, we believe that the adoption of the RDF standards provide
novel use cases with the ChEMBL data. More importantly, the adoption of explicit
semantics simplifies integration to other resources, and that is one of the requirements
core to the efforts of cheminformatics.

>>>>

Much could be argued about the choice of ontologies, and the lack of an
ontology in places where we decided to use custom classes and predicates. Indeed, the
choices have only partly been decided using ontological arguments. Most of the choices are based
on pragmatic approaches to enable solutions to the range of use cases we describe here. It should be noted,
however, that the ChEMBL database captures a wide variety of experimental studies
and that semantics are not fully specified in the primary source from which the ChEMBL data was
extracted. As it is not possible to infer the original intentions in these cases, 
it is our opinion that it is better to leave some semantics undefined. During the 
development of the applications some incorrect assumptions have been discovered and
fixed, but in the main we have tried to avoid making assumptions if they are not generally known to be valid. 
Therefore, the goal of this work is the conversion of the ChEMBL data into
RDF, and not an ontological modeling of the domain.

Once case where Linked Data publishers must avoid implying strong semantics in many cases is
in relation to links to other databases. When linking between two databases based on a shared identifier,
Linked Data publishers make assumptions including syntax assumptions that the identifiers 
are correct and present in the target database, along with semantic assumptions inferring that the
that entries identified in the target database have a scientifically meaningful relationship with the current database.
It is for this reason that the ChEMBL-RDF dataset uses different predicates for linking to various databases.
This problem, however, is general, applying also to other cases where databases link out to third-party sources using identifiers~\cite{VanIersel2010}.

Also debatable is the choice of ontologies used to describe the dataset in RDF. As an example, in this release of the
ChEMBL-RDF data, we adopted the cheminformatics ontology, CHEMINF. Using this ontology
we can encode numeric and non-numeric descriptors of chemical entities in a consistent
manner. This method is used to consistently expose identity descriptors such
as the SMILES string, InChI, and InChIKey, and molecular properties including the
logP, rotatable bond count, and hydrogen bond donor and acceptor counts.
However, the design of this annotation is more general than may be needed for a translation
of the data in the ChEMBL relational database into RDF. Although this results in more 
complex SPARQL queries, it allows generic queries that may not otherwise be easy to construct.
While the selection of ontologies used was found to fit well with our applications,
we anticipate that we will select more expressive ontologies in future iterations of the project.

Some information in the databases has not yet been annotated using standard ontologies. 
For example, the activity types in ChEMBL, ``IC50'', ``EC50'', and ``Potency'', are not exposed in the RDF dataset
meaning that it is not yet possible to reason or group over similar activity types.
When the activity types are exposed queries could be constructed based on the knowledge 
that the IC$_{50}$ and K$_i$ end points are closely related. Similarly, measurement units are currently provided as
string literals, making it difficult to compare concentrations when some are given in nM and others in mM.
In the future it may also be possible, relying on some assumptions, to crossmatch activities
expressed in $\mu$g/ml with query concentrations in nM. These additional semantics are necessary 
to translate raw experimental data into semantically useful, machine-readable information.

Two ontologies were recently developed that may address these issues for ChEMBL-RDF in future. Firstly, the BioAssay Ontology, which addresses many of our
remaining concerns around making the experimental data machine readable~\cite{Visser2011}. Secondly,
the Quantity, Unit, Dimension and Type (QUDT) ontology which makes it possible
to express measurements in a standardised machine readable way~\cite{QUDT}. Both approaches are 
being explored in the Open PHACTS project and will be applied to ChEMBL-RDF as the Open PHACTS project progresses~\cite{Williams2012}.

Taking these points into account, we believe that the adoption of RDF and community based ontologies allows
novel use cases to be solved using the ChEMBL data that would not otherwise be achievable. More importantly, the adoption of explicit
semantics simplifies integration to other resources, as others can more easily understand the intent and purpose of each part of the ChEMBL dataset.
This is a requirement which is essential to the efforts of cheminformatics researchers wishing to infer knowledge from both ChEMBL and other datasets.
